---
title: "NYC Emergency Medical Visits"
author: "Louie Sui"
date: "2022-11-02"
output: pdf_document
---

(The dataset is a log of the number of emergency department visit. Each record captures the number of emergency visits in a given time frame.)

# Import Libraries
These are the packages that we will use in our analysis. All of which contains the necessary functions we will use. 
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(rvest)
library(sf)
library(tidyverse)
library(jsonlite)
library(tidycensus)
```

# Loading Emergency Visit Data
For this analysis, we will use a dataset containing a log of the number of different visits in each zip codes in New York. The dataset will be obtained from the "cityofnewyork.us". The cityofnewyork.us is a reliable source for public datasets. Datasets from this source are managed and maintained by data coordinators. Thus, this makes our data more accurate.    
```{r}
emergencyVisit =
  fromJSON("https://data.cityofnewyork.us/resource/2nwg-uqyg.json?$limit=3000")
```

# Cleaning the Data
#### Convert "date" to date format
After importing the dataset, we find that the data types of the columns imported as *char*. We will convert the columns to their appropriate data types. This allows us to aggregate the dataset later.
```{r}
emergencyVisit$date = as.Date(emergencyVisit$extract_date)
emergencyVisit$extract_date = as.Date(emergencyVisit$extract_date)

emergencyVisit$total_ed_visits = strtoi(emergencyVisit$total_ed_visits)
```

#### Filtering to recent 300 records
For our analysis, we only analyze a small portion of the data. We save the new dataset into a new dataframe called, "recentVisit". 
```{r}
recentVisit <- emergencyVisit %>% 
  arrange(desc(date)) %>% 
  head(300)
```

# Data Exploration
#### Calculate the number of visits per ZIP code
After cleaning the dataset, we aggregated the dataset using the max function on the number of visits received in each neighborhood. We find that the neighborhood with the most amount of visits is 10456. 
```{r}
recentVisit %>% 
  group_by(mod_zcta) %>% 
  summarize(totalVisit=max(total_ed_visits)) %>% 
  arrange(desc(totalVisit))
```

# PART 2
#### Getting Geolocation Data
In this part, we import a geolocation dataset. We will use the geolocation dataset to visualize our findings to our stakeholders. This will make it easy for our stakeholders to understand our analysis. 
```{r}
nycMap = st_read("/Users/ls/Documents/cis 9750/Assignment 3/ZIP_CODE_040114/ZIP_CODE_040114.shp")
```

#### Graph a map outlining the ZIP codes of NYC. 
The following code maps the borders to the map we will use. We can confirm that this map is correct by confirming that it is identical to New York's borders. 
```{r}
plot(st_geometry(nycMap))
```

#### Merge recentVisit and geolocation datasets
We merge the datasets together using the merge function. 
```{r}
nycMapGeo <- nycMap %>% 
  mutate(mod_zcta=as.character(ZIPCODE)) %>% 
  select(mod_zcta, geometry)

geo_visit = merge(x=nycMapGeo, y=recentVisit, by='mod_zcta', all=TRUE)
```

#### Shade the map based on the number of visits in each ZIP code
The brighter colored zip codes indicates higher numbers of emergency visit and the darker, more blue, colored zip codes indicate lower total number of emergency visits. In the map below, we can see that areas in Brooklyn and Manhattan are more brightly lit than other areas. This means that these areas have the most amount of emergency visits in New York. However, using this metric alone to base our analysis is dangerous. We would need to dive deeper to see the entire picture. 
```{r}
total_visits <- geo_visit %>% 
  group_by(mod_zcta) %>% 
  summarise(total_ed_visits_2 = sum(total_ed_visits))
  
plot(geo_visit['total_ed_visits'])
```

# Part 3
#### Get Census Data
In this part, we will further our analysis by incorporating population data to our dataset. Simply just having the total number of emergency visits will not allow us to create a deep analysis of the dataset. The population dataset gives context to the number of emergency visits. Some zip codes which have an extremely high number of emergency visits is due to a high population count. Adding population of each zip code will allow us to make a more accurate judgement. 
```{r message=FALSE}
census_api_key("aa512886c5449a582d837da8d3a07af66a043fe5", install=TRUE, overwrite=TRUE)

censusData = load_variables(2018, "acs5", cache=T)
```

#### Download population data from the census
```{r message=FALSE}
populationData = get_acs(geography = "zcta"
                         , variables = 'B01003_001'
                         , geometry = FALSE)
```
 
#### Remove the spurious "ZCTA5" from the front of ZIP codes
```{r}
populationData$mod_zcta = sub("ZCTA5 ", "", populationData$NAME)
```

#### Rename the population column
```{r}
populationData <- populationData %>% 
  rename(population=estimate)
```

#### Merge the population data with geo_visits dataset
```{r}
geo_visit = merge(x=geo_visit, y=populationData, by='mod_zcta', all.x=TRUE)
```

#### Create a column that reports the rate of Emergency visits per inhabitants
In this column, we can see that the average rate of emergency visit is 10.87%. This means that on average about 1 in 10 inhabitants in New York are projected to visit the emergency room. However, there are are also some locations where the rate is as low as 0.4%. Whether this means that inhabitants are extremely cautious or inhabitants are averse to going to emergency room still requires further analysis. 
```{r}
geo_visit <- geo_visit%>% 
  mutate(rate_ed_visit=total_ed_visits/population * 100) %>% 
  arrange(desc(rate_ed_visit))

summary(geo_visit)
```

# Part 4
#### Download Poverty Data
Next, we test the hypothesis of poverty on the total number of ed visits. It seems likely that the higher the poverty rate, the higher the number of ED visits would be. 
```{r}
vars = c(poverty = 'B17001_002')

povertyData = get_acs(geography = "zcta",
                      variables = vars,
                      geometry = FALSE)
```

#### Remove the spurious "zcta5" from the ZIP data and clean up 
```{r}
povertyData$mod_zcta = sub("ZCTA5 ", "", povertyData$NAME)

povertyData = povertyData %>% rename(povertyCount = estimate)
```

#### Remove unneccessary columns in our dataset. 
```{r}
povertyData = povertyData %>% 
  select(-c(GEOID, NAME, variable, moe))
```

#### Perform an inner join of the dataframe you had with the poverty data just downloaded
```{r}
geo_visit <- merge(x=geo_visit, y=povertyData, by="mod_zcta")
```

#### Calculate the poverty rate
```{r}
geo_visit <- geo_visit %>% 
  mutate(poverty_rate = povertyCount/population * 100)
```

#### Analyze the linear relationship between poverty rate and the number of visits, across ZIP codes.
In the graph below, we can see that our hypothesis is correct. There is a positive relationship between poverty rate and the total number of ED visits. As poverty rate rises, the total number of emergency visits are also higher. This is not to say that higher poverty levels causes more ED visits but just to state that higher poverty rate tends to also have higher ED visits.

However, another observation of the relationship between the number of ED visits and the poverty rate is that as the poverty rate increases, the spread of the total number of ED visits become greater. It becomes more and more uncertain whether a ZIP code with higher poverty rate will have high number of ED visits or lower number of ED visits. This may be due to other factors such as patients in particular areas are more resistant to the emergency department or cost-saving initiatives  incorporated by patients. The exact reason requires more data. 
```{r}
ggplot(data=geo_visit, aes(x=poverty_rate, y=total_ed_visits)) +
  geom_point() +
  geom_smooth(method="lm", formula="y~x") +
  xlim(0,40)
```

#### Analyze the statistical significance of the regression.
According to the model that we produced, poverty rate is a significant factor in determining the number of ED visits. Despite the regression having a low R-squared value of only .4312, this model gives us a general idea of the impact poverty rate may have on the number of ED visits. For instance, if poverty rate increases by 1 percent, the number of ED visits would also increase between 2.8 to 3.2 more visits. 
```{r}
poverty_visit = lm(total_ed_visits~poverty_rate, data=geo_visit)

summary(poverty_visit)
```

#### Analyze the linear relationship between poverty rate and the rate of visits per inhabitants, across ZIP codes.
Like the previous model, there is a similar relationship between poverty rate and the rate of ED visits. As the poverty rate increases, the rate of ED visits would also increase. In other words, ZIP codes with higher level of poverty rate are expected to also have higher rate of ED visits. 

Lastly, the model also has wider spread. Locations with higher poverty rates have a wider range of ED rates. 
```{r}
ggplot(data=geo_visit, aes(x=poverty_rate, y=rate_ed_visit)) +
  geom_point() +
  stat_smooth(method="lm", col="red", formula="y~x") +
  xlim(0, 40)
```

#### Analyze the statistical significance of the linear regression. 
Analyzing the model, we see that the variables are statistically significant. This means that there is enough evidence to say that the poverty rate is significant in determining the rate of ed visits. However, just like the previous model, it is just as unreliable to predict the precise rate of ED visits using just the poverty rate. We can only get the idea that higher poverty rate tend to have higher ED rates. 
```{r}
poverty_visit_rate <- lm(rate_ed_visit~poverty_rate, data=geo_visit)

summary(poverty_visit_rate)
```

# Part 5
#### Download the age data
In this part, we will test the hypothesis of age on the total number of ed visits. We predict that the older the average individual in a location, the higher the total number of ed visits will be. 
```{r}
vars = c(age = 'B01002_001')
ageData = get_acs(geography = "zcta", variables = vars, geometry = FALSE)
```

#### Remove "ZCTA5" from the ZIP codes
```{r}
ageData$mod_zcta = sub("ZCTA5 ","",ageData$NAME)
```

#### Merge the data with the main dataset
```{r}
ageData <- ageData %>% 
  rename(age='estimate') %>% 
  select(age, mod_zcta)

geo_visit = merge(x=geo_visit, y=ageData, by="mod_zcta")
```

#### Analyze the relationship between age and ED visits
We can see from the graph below that age and ED visits are negatively correlated. This is to say that as age increases, the number of ED visits decreases. This is rather surprising as one would typically expect that as a person ages, patients would have more visits to the hospital. Perhaps, older individuals are less likely to engage in risky activities that would send them to the emergency department than younger people. 

On the other hand, as age increases, the number of ED visits a location has becomes more predictable. In fact, as age increases, the number of ed visits approaches zero. 
```{r}
age_visit_model <- lm(total_ed_visits~age, data=geo_visit)

summary(age_visit_model)

ggplot(data=geo_visit, aes(x=age, y=total_ed_visits)) +
  geom_point() +
  geom_smooth(method="lm", col="red", formula="y~x") +
  xlim(30,50)
```

# Part 6
#### Download the data from census bureau
Last but not least, we study the effect that white only demographics have on total number of ed visits. 
```{r}
vars = c('B02001_002')
ethnicityData = get_acs(geography = "zcta", variables = vars, geometry = FALSE)
ethnicityData = ethnicityData %>% rename(whiteAloneCount = estimate)
ethnicityData$mod_zcta = sub("ZCTA5 ","",ethnicityData$NAME)
```

#### Merge the ethnicity data into the main data set
```{r}
ethnicityData <- ethnicityData %>% 
  select(whiteAloneCount, mod_zcta)

geo_visit <- merge(x=geo_visit, y=ethnicityData, by="mod_zcta")
```

#### Calculate the percentage of the population that states they are "White Alone"
```{r}
geo_visit <- geo_visit %>% 
  mutate(white_ratio=whiteAloneCount/population * 100)
```

#### Perform a multi-linear regression of ED visits per inhabitant (for each ZIP) against 3 independent variables: “White Alone” percentage, median age, and poverty rate
```{r}
visit_multi_model <- lm(total_ed_visits ~ white_ratio + age + poverty_rate, data=geo_visit)
summary(visit_multi_model)
```

#### Analyze the strength of the multi-linear relationship
In the above model, we obtain information about each variable. For one, we can see that the intercept is 98% significant. We are 98% confident that the intercept is other than zero. In this case, when all other variables equal zero, there is about 68 ED visits. Furthermore, we can see that white ratio and poverty rate are also at high confidence levels. Both of these variable play a significant role in the number of ED visits. Therefore, white_ratio and poverty rate are important in predicting the number of ED visits observed. On the other hand, the regression model shows that age does not play a significant role in predicting the ED visits. We can remove age from the model to get a more precise measure of the impact of the other variables on the number of ED visits.

#### Perform multi-regression model without age
```{r}
visit_multi_model <- lm(total_ed_visits ~ white_ratio + poverty_rate, data=geo_visit)
summary(visit_multi_model)
```

#### Analysis of the relationship between white ratio and poverty rate
After removing age from our model, we get our final model. If the number of white_ratio and poverty_rate are zero, we can expect that there are about 33 visits. Meanwhile, when the white ratio increases by a factor of 1, we see a .406 drop in number of ED visits up to a maximum drop of 40.6. However, only some areas in New York contain a 100% white population. Therefore, the max number sound not be regarded as normal. Lastly, as poverty rate increases by a factor of 1, the number of ED visits will increase by 2.4131 up to a maximum increase of 172 ED visits. Poverty rate of locations in New York only goes up to 72%. 


